---
title: "Data Exploration"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r load Data, warning = F, message = F}
library(dplyr)
library(tidyr)
library(qqman)
library(gldrm)
library(ggplot2)
library(ggfortify)
setwd("/gpfs/ts0/projects/Research_Project-191406/Aisha/data/Array")
load("/gpfs/ts0/projects/Research_Project-191406/Aisha/data/Array/Normalised_Data_Sesame.rdat")


pheno <- QCmetrics
betas <- Normalised_Sesame_Betas

pheno_ecx <- pheno[which(pheno$Tissue %in% "CortexEntorhinalis"),]
betas_ecx <- betas[, colnames(betas) %in% pheno_ecx$Basename]
pheno_ecx_J20 <- pheno_ecx[which(pheno_ecx$AD_model %in% "J20"),]
betas_ecx_J20 <- betas_ecx[, colnames(betas_ecx) %in% pheno_ecx_J20$Basename]
pheno_ecx_Tg4510 <- pheno_ecx[which(pheno_ecx$AD_model %in% "rTg4510"),]
betas_ecx_Tg4510 <- betas_ecx[, colnames(betas_ecx) %in% pheno_ecx_Tg4510$Basename]
```

## PCA

```{r pca, warning = F, message = F}
# remove the sorted samples
pheno <- pheno[which(pheno$Tissue %in% c("CortexEntorhinalis","Hippocampus")),]
betas <- betas[, colnames(betas) %in% pheno$Basename]

pheno <- pheno[order(rownames(pheno)),]
betas <- betas[,order(colnames(betas))]
print(identical(rownames(pheno), colnames(betas)))

autoplot(prcomp(t(betas)), data = pheno, colour = 'Tissue', shape = 'AD_model')
```

#### rTg4510

```{r pca rTg , warning = F, message = F}
pheno_rTg <- pheno[which(pheno$AD_model == "rTg4510"),]
betas_rTg <- betas[,colnames(betas) %in% pheno_rTg$Basename]

pheno_rTg <- pheno_rTg[order(rownames(pheno_rTg)),]
betas_rTg <- betas_rTg[,order(colnames(betas_rTg))]
print(identical(rownames(pheno_rTg), colnames(betas_rTg)))

autoplot(prcomp(t(betas_rTg)), data = pheno_rTg, colour = 'Genotype', shape = 'Tissue')
```


#### J20


```{r pca J20 , warning = F, message = F}
pheno_J20 <- pheno[which(pheno$AD_model == "J20"),]
betas_J20 <- betas[,colnames(betas) %in% pheno_J20$Basename]

pheno_J20 <- pheno_J20[order(rownames(pheno_J20)),]
betas_J20 <- betas_J20[,order(colnames(betas_J20))]
print(identical(rownames(pheno_J20), colnames(betas_J20)))

autoplot(prcomp(t(betas_J20)), data = pheno_J20, colour = 'Genotype', shape = 'Tissue')
```

## Hierarchal clustering

```{r clsutering, warning = F, message = F}

library(gplots)
sigma<-apply(betas, 1, sd)# this is calcualtion the standard deviation of all probes

plot(hclust(dist(t(betas[order(sigma, decreasing = TRUE)[1:5000],]))), 
     labels = pheno$Tissue, hang = -1 , cex = 0.68, main = "", xlab = "Tissue") 

plot(hclust(dist(t(betas[order(sigma, decreasing = TRUE)[1:5000],]))), 
     labels = pheno$AD_model, hang = -1 , cex = 0.68, main = "", xlab = "AD_model") 
```

## Heatmaps

Heatmap is made by plotting top 1000 probes with the highest variance of all 23,633 probes. Interesting there is seperation of tissue and AD model but not genotype.

```{r heatmaps, warning = F, message = F, fig.height=8}
pheno$Chip_ID <- pheno$Slide

#All samples
cell_tissue <- as.data.frame(pheno$Tissue, rownames(pheno))
colnames(cell_tissue)<- "Tissue"
cell_age <- as.data.frame(pheno$AgeDays, rownames(pheno))
colnames(cell_age)<- "Age"
cell_model <- as.data.frame(pheno$AD_model, rownames(pheno))
colnames(cell_model)<- "AD_model"
cell_genotype <- as.data.frame(pheno$Genotype, rownames(pheno))
colnames(cell_genotype)<- "Genotype"
cell_chip<- as.data.frame(pheno$Slide, rownames(pheno))
colnames(cell_chip)<- "Chip"

metadata <- cbind(cell_age,cell_genotype,cell_model,cell_tissue, cell_chip)

library(pheatmap)
pheatmap(betas[order(sigma, decreasing = TRUE)[1:1000],],          
         annotation_col = metadata,
         cluster_rows = F,
         show_rownames = F,
         show_colnames = F,
         main = "All samples")


# ECX samples
cell_tissue <- as.data.frame(pheno_ecx$Tissue, rownames(pheno_ecx))
colnames(cell_tissue)<- "Tissue"
cell_age <- as.data.frame(pheno_ecx$AgeDays, rownames(pheno_ecx))
colnames(cell_age)<- "Age"
cell_model <- as.data.frame(pheno_ecx$AD_model, rownames(pheno_ecx))
colnames(cell_model)<- "AD_model"
cell_genotype <- as.data.frame(pheno_ecx$Genotype, rownames(pheno_ecx))
colnames(cell_genotype)<- "Genotype"
cell_chip<- as.data.frame(pheno_ecx$Slide, rownames(pheno_ecx))
colnames(cell_chip)<- "Chip"

metadata <- cbind(cell_age,cell_genotype,cell_model,cell_tissue, cell_chip)

library(pheatmap)
sigma<-apply(betas_ecx, 1, sd)
pheatmap(betas_ecx[order(sigma, decreasing = TRUE)[1:1000],],          
         annotation_col = metadata,
         cluster_rows = F,
         show_rownames = F,
         show_colnames = F, 
         main = "ECX")

# ECX samples J20s
cell_tissue <- as.data.frame(pheno_ecx_J20$Tissue, rownames(pheno_ecx_J20))
colnames(cell_tissue)<- "Tissue"
cell_age <- as.data.frame(pheno_ecx_J20$AgeDays, rownames(pheno_ecx_J20))
colnames(cell_age)<- "Age"
cell_model <- as.data.frame(pheno_ecx_J20$AD_model, rownames(pheno_ecx_J20))
colnames(cell_model)<- "AD_model"
cell_genotype <- as.data.frame(pheno_ecx_J20$Genotype, rownames(pheno_ecx_J20))
colnames(cell_genotype)<- "Genotype"
cell_chip<- as.data.frame(pheno_ecx_J20$Slide, rownames(pheno_ecx_J20))
colnames(cell_chip)<- "Chip"

metadata <- cbind(cell_age,cell_genotype,cell_model,cell_tissue, cell_chip)


sigma<-apply(betas_ecx_J20, 1, sd)
pheatmap(betas_ecx_J20[order(sigma, decreasing = TRUE)[1:1000],],          
         annotation_col = metadata,
         cluster_rows = F,
         show_rownames = F,
         show_colnames = F,
         main = "ECX J20")

# ECX samples Tg4510
cell_tissue <- as.data.frame(pheno_ecx_Tg4510$Tissue, rownames(pheno_ecx_Tg4510))
colnames(cell_tissue)<- "Tissue"
cell_age <- as.data.frame(pheno_ecx_Tg4510$AgeDays, rownames(pheno_ecx_Tg4510))
colnames(cell_age)<- "Age"
cell_model <- as.data.frame(pheno_ecx_Tg4510$AD_model, rownames(pheno_ecx_Tg4510))
colnames(cell_model)<- "AD_model"
cell_genotype <- as.data.frame(pheno_ecx_Tg4510$Genotype, rownames(pheno_ecx_Tg4510))
colnames(cell_genotype)<- "Genotype"
cell_chip<- as.data.frame(pheno_ecx_Tg4510$Slide, rownames(pheno_ecx_Tg4510))
colnames(cell_chip)<- "Chip"

metadata <- cbind(cell_age,cell_genotype,cell_model,cell_tissue, cell_chip)


sigma<-apply(betas_ecx_Tg4510, 1, sd)
pheatmap(betas_ecx_Tg4510[order(sigma, decreasing = TRUE)[1:1000],],          
         annotation_col = metadata,
         cluster_rows = F,
         show_rownames = F,
         show_colnames = F,
         main = "ECX rTg4510")


```

## Mouse Model Correlations

#### Probes

We can check that the methyaltion of the arrays are similair between similiar biological groups.All levels will be compared, e.g AD model and genotype

```{r probe correlation, warning = F, message = F, fig.show = "hold", out.width = "50%"}
#####################  ECX  ##################

pheno_ecx_J20 <- pheno_ecx[which(pheno_ecx$AD_model %in% "J20"),]
betas_ecx_J20 <- betas_ecx[, colnames(betas_ecx) %in% pheno_ecx_J20$Basename]

pheno_ecx_Tg4510 <- pheno_ecx[which(pheno_ecx$AD_model %in% "rTg4510"),]
betas_ecx_Tg4510 <- betas_ecx[, colnames(betas_ecx) %in% pheno_ecx_Tg4510$Basename]


# Calculte average methylation for all ind for each probe ECX
J20_probes <- apply(betas_ecx_J20,1,mean)
Tg4510_prones <- apply(betas_ecx_Tg4510, 1, mean)

plot(J20_probes, Tg4510_prones, main = "J20 rTg4510 ECX")

################### HIP ################


pheno_hip <- pheno[which(pheno$Tissue %in% "Hippocampus"),]
betas_hip <- betas[, colnames(betas) %in% pheno_hip$Basename]


pheno_hip_J20 <- pheno_hip[which(pheno_hip$AD_model %in% "J20"),]
betas_hip_J20 <- betas_hip[, colnames(betas_hip) %in% pheno_hip_J20$Basename]

pheno_hip_Tg4510 <- pheno_hip[which(pheno_hip$AD_model %in% "rTg4510"),]
betas_hip_Tg4510 <- betas_hip[, colnames(betas_hip) %in% pheno_hip_Tg4510$Basename]


# Calculte average methylation for all ind for each probe hip
J20_probes <- apply(betas_hip_J20,1,mean)
Tg4510_prones <- apply(betas_hip_Tg4510, 1, mean)

# colours WT TG

plot(J20_probes, Tg4510_prones, main = "J20 rTg4510 HIP", col = )


############# ECX J20 Genotypes ##########

pheno_ecx_J20_WT <- pheno_ecx_J20[which(pheno_ecx_J20$Genotype %in% "WT"),]
betas_ecx_J20_WT <- betas_ecx_J20[, colnames(betas_ecx_J20) %in% pheno_ecx_J20_WT$Basename]

pheno_ecx_J20_TG <- pheno_ecx_J20[which(pheno_ecx_J20$Genotype %in% "TG"),]
betas_ecx_J20_TG <- betas_ecx_J20[, colnames(betas_ecx_J20) %in% pheno_ecx_J20_TG$Basename]

ECX_J20_WT <- apply(betas_ecx_J20_WT,1,mean)
ECX_J20_TG <- apply(betas_ecx_J20_TG, 1, mean)

plot(ECX_J20_WT,ECX_J20_TG, main = "J20 WT vs TG ECX")

############# ECX Tg4510 Genotypes ##########

pheno_ecx_Tg4510_WT <- pheno_ecx_Tg4510[which(pheno_ecx_Tg4510$Genotype %in% "WT"),]
betas_ecx_Tg4510_WT <- betas_ecx_Tg4510[, colnames(betas_ecx_Tg4510) %in% pheno_ecx_Tg4510_WT$Basename]

pheno_ecx_Tg4510_TG <- pheno_ecx_Tg4510[which(pheno_ecx_Tg4510$Genotype %in% "TG"),]
betas_ecx_Tg4510_TG <- betas_ecx_Tg4510[, colnames(betas_ecx_Tg4510) %in% pheno_ecx_Tg4510_TG$Basename]

ECX_Tg4510_WT <- apply(betas_ecx_Tg4510_WT,1,mean)
ECX_Tg4510_TG <- apply(betas_ecx_Tg4510_TG, 1, mean)

plot(ECX_Tg4510_WT,ECX_Tg4510_TG, main = "Tg4510 WT vs TG ECX")

############# hip J20 Genotypes ##########

pheno_hip_J20_WT <- pheno_hip_J20[which(pheno_hip_J20$Genotype %in% "WT"),]
betas_hip_J20_WT <- betas_hip_J20[, colnames(betas_hip_J20) %in% pheno_hip_J20_WT$Basename]

pheno_hip_J20_TG <- pheno_hip_J20[which(pheno_hip_J20$Genotype %in% "TG"),]
betas_hip_J20_TG <- betas_hip_J20[, colnames(betas_hip_J20) %in% pheno_hip_J20_TG$Basename]

hip_J20_WT <- apply(betas_hip_J20_WT,1,mean)
hip_J20_TG <- apply(betas_hip_J20_TG, 1, mean)

plot(hip_J20_WT,hip_J20_TG, main = "J20 WT vs TG hip")


############# hip Tg4510 Genotypes ##########

pheno_hip_Tg4510_WT <- pheno_hip_Tg4510[which(pheno_hip_Tg4510$Genotype %in% "WT"),]
betas_hip_Tg4510_WT <- betas_hip_Tg4510[, colnames(betas_hip_Tg4510) %in% pheno_hip_Tg4510_WT$Basename]

pheno_hip_Tg4510_TG <- pheno_hip_Tg4510[which(pheno_hip_Tg4510$Genotype %in% "TG"),]
betas_hip_Tg4510_TG <- betas_hip_Tg4510[, colnames(betas_hip_Tg4510) %in% pheno_hip_Tg4510_TG$Basename]

hip_Tg4510_WT <- apply(betas_hip_Tg4510_WT,1,mean)
hip_Tg4510_TG <- apply(betas_hip_Tg4510_TG, 1, mean)

plot(hip_Tg4510_WT,hip_Tg4510_TG, main = "Tg4510 WT vs TG hip")

```


## Pathology

There are pathylogy data for the two tissues. This is based off the bumber of antibodies staied to the protein of interst. The times of this assay differs for both models so we will look at the association of pathology with age seperately for the models.


```{r pathology HIP, warning = F, message = F,fig.show='hold',out.weight="50%"}
library(ggpubr)
par(mfrow = c(1,2))
#############   Join the pathology dataset with the pheno file  ###################

# load("../Normalised_Data_Sesame.rdat")
dir_path <- "/gpfs/ts0/projects/Research_Project-191406/Aisha/data/Array/"
J20_ECX <- read.csv(paste(dir_path, "J20_ECX_Pathology.csv", sep = ""), stringsAsFactors = F, header = T)
J20_HIP <- read.csv(paste(dir_path, "J20_HIP_Pathology.csv", sep = ""), stringsAsFactors = F, header = T)
Tg4510_ECX <- read.csv(paste(dir_path, "Tg4510_ECX_Pathology.csv", sep = ""), stringsAsFactors = F, header = T)
Tg4510_HIP <- read.csv(paste(dir_path, "Tg4510_HIP_Pathology.csv", sep = ""), stringsAsFactors = F, header = T)


pheno$Basename <- paste(pheno$Slide, pheno$stripe, sep = "_")
rownames(pheno) <- pheno$Basename


# Add mouse model to external sample name
J20_ECX$ExternalSampleID <- paste("J20_", J20_ECX[,1], sep = "")
J20_HIP$ExternalSampleID <- paste("J20_", J20_HIP[,1], sep  ="")
J20_ECX <- J20_ECX[,2:9]
J20_HIP <- J20_HIP[,2:9]
J20 <- rbind(J20_ECX, J20_HIP)

Tg4510_ECX$ExternalSampleID <- paste("rTg4510_", Tg4510_ECX[,1], sep = "")
Tg4510_HIP$ExternalSampleID <- paste("rTg4510_", Tg4510_HIP[,1], sep  ="")
Tg4510_ECX <- Tg4510_ECX[,2:9]
Tg4510_HIP <- Tg4510_HIP[,2:9]
Tg4510 <- rbind(Tg4510_ECX, Tg4510_HIP)

path <- rbind(J20, Tg4510)
pheno <- merge(path, pheno, by = "ExternalSampleID")


# Tissue, AD_model 1 -  Cortex J20
pheno_ecx_J20 <- pheno[which(pheno$Tissue == "CortexEntorhinalis" & pheno$AD_model == "J20"),]
ggplot(pheno_ecx_J20 , aes(x = pheno_ecx_J20$Age_months, y = pheno_ecx_J20$Pathology_ECX, 
                           color = pheno_ecx_J20$Genotype.x ))+
  geom_point() +
  geom_smooth(method=lm, se=FALSE) +
  theme_classic() +
  stat_cor(method="pearson") +
  labs(title = "ECX J20", x = "Age_months", y = "Pathology_ECX") +
  guides(color=guide_legend(title="Genotype"))

# Tissue, AD_model 2 - Corted Tg4510
pheno_ecx_Tg4510 <- pheno[which(pheno$Tissue == "CortexEntorhinalis" & pheno$AD_model == "rTg4510"),]
ggplot(pheno_ecx_Tg4510 , aes(x = pheno_ecx_Tg4510$Age_months, y = pheno_ecx_Tg4510$Pathology_ECX, 
                           color = pheno_ecx_Tg4510$Genotype.x ))+
  geom_point() +
  geom_smooth(method=lm, se=FALSE) +
  theme_classic() +
  stat_cor(method="pearson") +
  labs(title = "ECX Tg4510", x = "Age_months", y = "Pathology_ECX") +
  guides(color=guide_legend(title="Genotype"))

# Tissue, AD_model 3 - Hippo J20
pheno_hip_J20 <- pheno[which(pheno$predictedTissue == "Hippocampus" & pheno$AD_model == "J20"),]
ggplot(pheno_hip_J20 , aes(x = pheno_hip_J20$Age_months, y = pheno_hip_J20$Pathology_HIP, 
                           color = pheno_hip_J20$Genotype.x ))+
  geom_point() +
  geom_smooth(method=lm, se=FALSE) +
  theme_classic() +
  stat_cor(method="pearson") +
  labs(title = "HIP J20", x = "Age_months", y = "Pathology_HIP") +
  guides(color=guide_legend(title="Genotype"))

# Tissue, AD_model 4 - Hippo Tg4510
pheno_hip_Tg4510 <- pheno[which(pheno$predictedTissue == "Hippocampus" & pheno$AD_model == "rTg4510"),]
ggplot(pheno_hip_Tg4510 , aes(x = pheno_hip_Tg4510$Age_months, y = pheno_hip_Tg4510$Pathology_HIP, 
                              color = pheno_hip_Tg4510$Genotype.x ))+
  geom_point() +
  geom_smooth(method=lm, se=FALSE) +
  theme_classic() +
  stat_cor(method="pearson") +
  labs(title = "HIP Tg4510", x = "Age_months", y = "Pathology_HIP") +
  guides(color=guide_legend(title="Genotype"))



```