---
title: "RRBS - BiSeq J20 Part 2 - Adapted Beta Regression"
author: "Isabel Castanho (I.S.Castanho@exeter.ac.uk)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true # to create the table of contents
    toc_float: false # make it float so that as you scroll down the page, the toc will always be visible
    number_sections: false # number_sections is for automatic numbering of #, ## etc. sections
    code_folding: hide
---

```{css, echo=FALSE}
body .main-container {
  max-width: 1800px !important;
  margin-left: auto !important;
  margin-right: auto !important;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/')

# # # load packages
# # if (!requireNamespace("BiocManager", quietly = TRUE))
# #     install.packages("BiocManager")
# # 
# # BiocManager::install("BiSeq")
# 
library(BiSeq)
library(betareg)
library(gplots)
library(cgwtools) # package that allows to resave objects in R using resave(..., list = character(), file)
library(lmtest) # to use likelihood-ratio test (lrt)

# load(file = "/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/biseq_J20.RData") # complete file
load(file = "/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/biseq_predictedMeth_J20.RData") # predictedMeth
load(file = "/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/biseqBetaResults_J20.RData")

color_J20_TG <- "#FF5A62"

# source("/gpfs/ts0/projects/Research_Project-191406/Aisha/script/rrbs-ad-mice/betaRegressionAdapted.R")
```

To detect the CpG sites where the DNA methylation differs between TG and WT samples we model the methylation within a beta regression./

<br/>

```{r}
numCores <- detectCores()
numCores
```

### 3. Model and test group effect
```{r}
# colData(predictedMeth)$Age_months <- as.numeric(as.character(colData(predictedMeth)$Age_months))
# # Age had been saved as factor so we changed to numeric before running the beta regression
# 
# betaResults <- betaRegressionAdapted(formula = ~Genotype + Age_months + Genotype*Age_months,
#                                      formulaNull = ~Genotype + Age_months,
#                                      link = "probit",
#                                      object = predictedMeth,
#                                      type = "BR",
#                                      mc.cores = 16)
# # betaResults is a data.frame containing model and test information for each CpG site
# # By setting type = "BR" the maximum likelihood with bias reduction is called
# 
# save(betaResults, file = "/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/biseqBetaResults_J20.RData")

```

Head results
```{r}
head(betaResults)
```

Summary results
```{r}
summary(betaResults)
```


###  4.1. Test CpG clusters for differential methylation
The aim is to detect CpG clusters containing at least one differentially methylated location.
To do so the P values p from the Wald tests are transformed to Z scores: z= Φ−1(1−p), which are normally distributed under Null hypothesis (no group effect).
As cluster test statistic a standardized Z score average is used.  To estimate the standard deviation of the Z scores we have to estimate the correlation and hence the variogram of methylation between two  CpG  sites  within  a  cluster.   The  estimation  of  the  standard  deviation requires that the distribution of the Z scores follows a standard normal distribution.  However, if methylation in both groups differs for many CpG sites the density distribution of P values shows a peak near 0.  
To ensure that the P values are roughly uniformly distributed to get a variance of the Z scores that is Gaussian with variance 1 we recommend to estimate the variogram (and hence the correlation of Z scores) under the null hypothesis.  To do so we model the beta regression again for resampled data:
```{r}
# # Both resampled groups should have the same number of
# # cases and control samples:
# # Take bottom 16 rows of the ColData predictedMeth
# predictedMethNull <- predictedMeth[,46:61]
# # Age and Genotype selected are random to not group based on genotype or age - we selected 16 for 2 genotype, 4 time points and repeated twice to increase randomness
# colData(predictedMethNull)$group.null <- rep(c(1,2), each = 8)
# print(colData(predictedMethNull))
# # To shorten the run time, please set mc.cores, if possible!
# betaResultsNull <- betaRegression(formula = ~group.null,
#                                         link = "probit",
#                                         object = predictedMethNull,
#                                         type="BR",
#                                         mc.cores = 16)
# 
# vario <- makeVariogram(betaResultsNull)
# 
# save(betaResultsNull, vario, predictedMethNull,
#       file = "/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/biseqBetaResultsNull_J20.RData")

```

```{r}
# Load the previous chunk betaResultsNull rdat object. Also print the samples used to generate the betaResultsNull
load("/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/biseqBetaResultsNull_J20.RData")
print(colData(predictedMethNull))
```
We estimate the variogram for the Z scores obtained for the resampled data.  
Based  on  the  variogram  plot  we  evaluate  the sill  (usually  near  1)  of  the variogram and smooth the curve:
```{r fig5, fig.height = 5, fig.width = 5, fig.cap = "Figure 5:  Estimated variogram together with the smoothed curve"}
plot(vario$variogram$v, ylim = c(0, 1.5))
vario.sm <- smoothVariogram(vario, sill = 0.9) # The vario.sm object is a list of two: "variogram" "pValsList"
lines(vario.sm$variogram[,c("h", "v.sm")],col = "red", lwd = 1.5)
grid()
```

```{r}
# Save all ojects from betaResultsNull
resave(betaResultsNull, vario, vario.sm, predictedMethNull,
     file = "/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/biseqBetaResultsNull_J20.RData")
```

```{r}
head(vario.sm$pValsList[[1]])
```


```{r}
# Subset the betaresults for: Genotype,Age,Genotype*Age interaction
load("/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/biseqBetaResults_J20.RData")
# Change colnames of the betaresults dataframe

# Genotype
betaResultsGenotype <- betaResults[,c(1:9,17)]
colnames(betaResultsGenotype) <- colnames(betaResultsNull)

# Interaction (meth groups 1&2 and diff are WT vs TG)
betaResultsInteraction <- betaResults[,c(1:2,13,4:6,14:15,9,17)]
colnames(betaResultsInteraction) <- colnames(betaResultsNull)

```

## Genotype

```{r}
# We replace the pValsList object (which consists of the test results of the resampled data) by the test results of interest (for group effect)
## auxiliary object to get the pValsList for the test
## results of interest

vario.aux <- makeVariogram(betaResultsGenotype, make.variogram=FALSE)
vario.sm$pValsList <- vario.aux$pValsList
head(vario.sm$pValsList[[1]])
```

vario.sm contains the smoothed variogram under the Null hypothesis together with the P values (and Z scores) from the Wald test, that the group has no effect on methylation.  The correlation of the Z scores between two locations in a cluster can now be estimated:
```{r}
# Correlaton is based on z-scores
locCor <- estLocCor(vario.sm)
```

We test each CpG cluster for the presence of at least one differentially methylated  location  at q what  can  be  interpreted  as  the  size-weighted  FDR  on clusters:
```{r}
clusters.rej <- testClusters(locCor, FDR.cluster = 0.10)

clusters.rej$clusters.reject
```

###  4.2. Trim significant CpG clusters
We then trim the rejected CpG clusters that is to remove the not dierentially
methylated CpG sites at q1 what can be interpreted as the location-wise FDR

```{r}
clusters.trimmed <- trimClusters(clusters.rej, FDR.loc = 0.05)
head(clusters.trimmed)
# clusters.trimmed is a data.frame object containing all differentially methylated CpG sites. The p.li column contains the P values estimated in the cluster trimming step
```

### 5. Defnition of DMR boundaries
We can now define the boundaries of DMRs as rejected CpG sites within
which rejected CpG sites solely are located.  
Within the DMRs the distance between neighbored rejected CpG sites should not exceed max.dist base pairs (usually the same as for max.dist in clusterSites), otherwise, the DMR is splitted.  
DMRs are also splitted if the methylation diference switches from positive to negative, or vice versa, if diff.dir = TRUE.  
That way we ensure that within a DMR all CpG sites are hypermethylated, and
hypomethylated respectively.
```{r}
DMRs <- findDMRs(clusters.trimmed,
                 max.dist = 500,
                 diff.dir = TRUE)
DMRs
DMRsGenotype <- as.data.frame(DMRs)
write.csv(DMRsGenotype,
          file = "/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/DMRs/J20/DMRsGenotype.csv")

DMRsGenotype <- DMRs
save(betaResultsGenotype, DMRsGenotype,
     file = "/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/biseqResultsGenotype_J20.RData")
```

## Interaction

```{r}
# We need to reload  betaResulstNull Objects & repeat as genotype section chunks
load("/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/biseqBetaResults_J20.RData")

```

```{r}
# We replace the pValsList object (which consists of the test results of the resampled data) by the test results of interest (for group effect)
## auxiliary object to get the pValsList for the test
## results of interest

vario.aux <- makeVariogram(betaResultsInteraction, make.variogram=FALSE)
vario.sm$pValsList <- vario.aux$pValsList
head(vario.sm$pValsList[[1]])
```

vario.sm contains the smoothed variogram under the Null hypothesis together with the P values (and Z scores) from the Wald test, that the group has no effect on methylation.  The correlation of the Z scores between two locations in a cluster can now be estimated:
```{r}
# Correlaton is based on z-scores
locCor <- estLocCor(vario.sm)
```

We test each CpG cluster for the presence of at least one differentially methylated  location  at q what  can  be  interpreted  as  the  size-weighted  FDR  on clusters:
```{r}
clusters.rej <- testClusters(locCor, FDR.cluster = 0.10)

clusters.rej$clusters.reject
```

###  4.2. Trim significant CpG clusters
We then trim the rejected CpG clusters that is to remove the not dierentially
methylated CpG sites at q1 what can be interpreted as the location-wise FDR

```{r}
clusters.trimmed <- trimClusters(clusters.rej, FDR.loc = 0.05)
head(clusters.trimmed)
# clusters.trimmed is a data.frame object containing all differentially methylated CpG sites. The p.li column contains the P values estimated in the cluster trimming step
```

### 5. Defnition of DMR boundaries
We can now define the boundaries of DMRs as rejected CpG sites within
which rejected CpG sites solely are located.  
Within the DMRs the distance between neighbored rejected CpG sites should not exceed max.dist base pairs (usually the same as for max.dist in clusterSites), otherwise, the DMR is splitted.  
DMRs are also splitted if the methylation diference switches from positive to negative, or vice versa, if diff.dir = TRUE.  
That way we ensure that within a DMR all CpG sites are hypermethylated, and
hypomethylated respectively.
```{r}
DMRs <- findDMRs(clusters.trimmed,
                 max.dist = 500,
                 diff.dir = TRUE)
DMRs
DMRsInteraction <- as.data.frame(DMRs)
write.csv(DMRsInteraction,
          file = "/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/DMRs/J20/DMRsInteraction.csv")

DMRsInteraction <- DMRs
save(betaResultsInteraction, DMRsInteraction,
     file = "/gpfs/ts0/projects/Research_Project-191406/isabel/RRBS_new/biseqResultsInteraction_J20.RData")
```
